Recovery of deleted deep sequencing data sheds more light on the early Wuhan SARS-CoV-2 epidemic

### 2021-06-17 paper/paper.tex
some progress on Conclusion

\title{Recovery of a deleted deep sequencing data set sheds more light on the early Wuhan SARS-CoV-2 epidemic}

@@

But in the case of Wuhan, genomic epidemiology has also proven frustratingly inconclusive.
Some of the problem is simply limited data: despite the fact that Wuhan has advanced virology labs, there is only patchy sampling of SARS-CoV-2 sequences from the first months of the city's explosive outbreak.
Other than a set of multiply sequenced samples collected in late December of 2019 from a dozen patients connected to the Huanan Seafood Market~\citep{WHO2021origins}, just a handful of Wuhan sequences are available from before late January of 2020 (see analysis in this study below).
The paucity of sequences could be due in part to an order that unauthorized Chinese labs destroy all coronavirus samples from early in the outbreak, reportedly for ``laboratory biological safety'' reasons~\citep{pinghui2020SCMP}.

@@

\subsection{The data are sufficient to determine the viral sequence from the start of spike through the end of ORF10 for some samples}
The approach of \citet{wang2020medRxiv} involved sequencing PCR amplicons covering nucleotide sites 21,563 to 29,674 of the SARS-CoV-2 genome, which spans from the start of the spike gene to the end of ORF10.
The approach also involved sequencing a short amplicon generated by nested PCR that covered a fragment of ORF1ab spanning sites $\sim$15,080 to 15,550.
In this paper, I only analyze the region from spike through ORF10 because this is a much longer contiguous sequence and the amplicons were generated by conventional rather than nested PCR.
I slightly trimmed the region of interest to sites 21,570 to 29,550 because many samples had poor coverage at the termini of the region.

@@

I also determined the consensus sequence of the plasmid control used by \citet{wang2020medRxiv} from the recovered sequencing data, and found that it had mutations C28144T and G28085T relative to proCoV2, which means that in the region of interest this control matches Wuhan-Hu-1 with the addition of G28085T.
Since none of the viral samples in Table~\ref{tab:mutations} contain G28085T and the samples that prove most relevant below also lack C28144T (which is a frequent natural mutation among early Wuhan sequences), this fact indicates plasmid contamination did not afflict the viral samples in the deleted sequencing project.

@@

This conundrum is illustrated in Figure~\ref{fig:deltadist_RaTG13}, which plots the collection date of SARS-CoV-2 sequences in GISAID versus the relative number of mutational differences from RaTG13~\citep{zhou2020pneumonia}, which is the bat coronavirus with the highest full-genome sequence identity to SARS-CoV-2.
The earliest SARS-CoV-2 sequences were collected in Wuhan in December, but these sequences are more distant from RaTG13 than sequences collected in January from other locations in China or even other countries (Figure~\ref{fig:deltadist_RaTG13}).
The discrepancy is especially pronounced for early Wuhan sequences from patients who had visited the Huanan Seafood Market, which was reported as an initial epicenter of the outbreak and has been postulated to be a location of potential zoonosis~\citep{WHO2021origins}.
All sequences associated with the Huanan Seafood Market differ from RaTG13 by at least three more mutations than sequences subsequently collected at various other locations (Figure~\ref{fig:deltadist_RaTG13})---a fact that is difficult to reconcile with the idea that the market was the original location of spread of a bat coronavirus into humans.
Importantly, all these observations also hold true if SARS-CoV-2 is compared to other related bat coronaviruses~\citep{lytras2021exploring} such as RpYN06~\citep{zhou2021identification} or RmYN02~\citep{zhou2020novel} rather than RaTG13 (Figure~\ref{suppfig:deltadist_RpYN06_RmYN02}).

 \begin{figure*}
 \centerline{
 \includegraphics[width=0.31\linewidth, valign=t, clip=true, trim=0in 4.4in 0in 0in]{figures/tree_images/hCoV-19-USA-WA1-2020_RaTG13_without_deleted_seqs.pdf}
 \hspace{0.02\linewidth}
 \includegraphics[width=0.33\linewidth, valign=t, clip=true]{figures/tree_images/hCoV-19-Guangdong-HKU-SZ-002-2020_RaTG13_without_deleted_seqs.pdf}
  \hspace{0.02\linewidth}
 \includegraphics[width=0.31\linewidth, valign=t, clip=true, trim=0in 4.4in 0in 0in]{figures/tree_images/hCoV-19-Shandong-LY005-2-2020_RaTG13_without_deleted_seqs.pdf}
 }
 \caption{
 Phylogenetic trees of SARS-CoV-2 sequences in GISAID with multiple observations among viruses collected no later than January 31, 2020.
 The trees are identical except they are rooted to make the putative progenitor each of the three sequences with highest identity to the RaTG13 bat coronavirus outgroup.
Nodes are shown as pie charts with areas proportional to the number of observations of that sequence, and colored by where the viruses were collected.
The mutations on each branch are labeled, with mutations towards the nucleotide identity in the outgroup in purple.
The labels at the top of each tree give the first known virus identical to each putative progenitor, as well as mutations in that progenitor relative to proCoV2~\citep{kumar2021evolutionary} and Wuhan-Hu-1.
The monophyletic group containing C28144T is collapsed into a single node labeled ``clade B'' in concordance with the naming scheme of \citet{rambaut2020dynamic}; this clade contains Wuhan-Hu-1.
 Figure~\ref{suppfig:tree_RpYN06_RmYN02} shows that identical results are obtained if the outgroup is RpYN06 or RmYN02.
\label{fig:tree_RaTG13}
 }
 \end{figure*}

This conundrum can be visualized in a phylogenetic context by rooting a tree of early SARS-CoV-2 sequences so that the ``progenitor'' sequence is closest to the bat coronavirus outgroup.
If we limit the analysis to sequences with at least two observations among strains collected no later than January 2020, there are three ways to root the tree in this fashion since there are three different sequences equally close to the outgroup (Figure~\ref{fig:tree_RaTG13}, Figure~\ref{suppfig:tree_RpYN06_RmYN02}).
Importantly, none of these rootings place any Huanan Seafood Market viruses (or other Wuhan viruses from December 2019) in the progenitor node---and only one of the rootings has any virus from Wuhan in the progenitor node (in the leftmost tree in Figure~\ref{fig:tree_RaTG13}, the progenitor node contains Wuhan/0126-C13/2020, which was reportedly collected on January 26, 2020).
Therefore, inferences about the progenitor of SARS-CoV-2 based on comparison to related bat viruses are inconsistent with other evidence suggesting the progenitor is an early virus from Wuhan~\citep{pipes2021assessing}.

@@

\begin{figure}
\centering
\includegraphics[width=0.75\linewidth]{figures/deltadist_jitter.png}
\caption{
Relative mutational distance from RaTG13 bat coronavirus outgroup calculated \emph{only} over the region of the SARS-CoV-2 genome covered by sequences from the deleted data set (sites 21,570--29,550).
The plot shows sequences in GISAID collected before February of 2020, as well as the 13 early Wuhan epidemic sequences in Table~\ref{tab:mutations}.
Mutational distance is calculated relative to proCoV2, and points are slightly jittered on the y-axis.
Go to {\color{red} add link} for an interactive version of this plot that enables toggling the outgroup to RpYN06 or RmYN02, mouseovers to see details for each point, and adjustment of y-axis jittering.
\label{fig:deltadist_jitter}
}
\end{figure}

 \begin{figure*}
 \centerline{
 \includegraphics[width=0.31\linewidth, valign=t, clip=true, trim=0in 6.75in 0in 0in]{figures/tree_images/hCoV-19-USA-WA1-2020_RaTG13_with_deleted_seqs.pdf}
 \hspace{0.02\linewidth}
 \includegraphics[width=0.33\linewidth, valign=t, clip=true]{figures/tree_images/hCoV-19-Guangdong-HKU-SZ-002-2020_RaTG13_with_deleted_seqs.pdf}
  \hspace{0.02\linewidth}
 \includegraphics[width=0.31\linewidth, valign=t, clip=true, trim=0in 6.75in 0in 0in]{figures/tree_images/hCoV-19-Shandong-LY005-2-2020_RaTG13_with_deleted_seqs.pdf}
 }
 \caption{
Phylogenetic trees like those in Figure~\ref{fig:tree_RaTG13} with the addition of the early Wuhan epidemic sequences from the deleted data set, and Guangdong patients infected in Wuhan prior to January 5 annotated separately.
Because the deleted sequences are partial, they cannot all be placed unambiguously on the tree.
Therefore, they are added to each compatible node proportional to the number of sequences already in that node.
The deleted sequences with C28144T (clade B) or C29095T (putative progenitor in middle tree) can be placed relatively unambiguously as defining mutations occur in the sequenced region, but those that lack either of these mutations are compatible with a large number of nodes including the proCoV2 putative progenitor.
 Figure~\ref{suppfig:tree_RpYN06_RmYN02} has already demonstrated that the results are identical if RpYN06 or RmYN02 is instead used as the outgroup.
\label{fig:tree_with_deleted}
 }
 \end{figure*}

Repeating the analysis of the previous section with these changes shows that several sequences from the deleted project and all sequences from patients infected in Wuhan but sequenced in Guangdong are more similar to the bat coronavirus outgroup than sequences from the Huanan Seafood Market (Figure~\ref{fig:deltadist_jitter}).
This fact suggests that the Huanan Seafood Market sequences, which are the primary focus of the genomic epidemiology in the joint WHO-China report~\citep{WHO2021origins}, may not be representative of the viruses that were circulating in Wuhan in late December of 2019 and early January of 2020.

@@

We can also qualitatively examine the three progenitor placements in Figure~\ref{fig:tree_with_deleted} using the principle employed by \citet{worobey2020emergence} to help evaluate scenarios for emergence of SARS-CoV-2 in Europe and North America: namely that during a growing outbreak, a progenitor is likely to give rise to multiple distinct branching lineages.
This principle is especially likely to hold for the scenarios in Figure~\ref{fig:tree_with_deleted}, since there are multiple individuals infected with each putative progenitor sequence, implying multiple opportunities to transmit descendants with new mutations.
Using this qualitative principle, the middle scenario in Figure~\ref{fig:tree_with_deleted} seems most plausible, the leftmost (proCoV2) scenario also seems plausible, and the rightmost scenario seems less plausible.
I acknowledge these arguments are purely qualitative and lack the formal statistical analysis of \citet{worobey2020emergence}---but as discussed below, there may be wisdom in qualitative reasoning when there are valid concerns about the nature of the underlying data.

\section{Discussion}

Something about lacking formal analyses of some other studies, but maybe this is not so bad.

>>>>

\title{Recovery of deleted deep sequencing data sheds more light on the early Wuhan SARS-CoV-2 epidemic}

@@

But in the case of Wuhan, genomic epidemiology has also proven frustratingly inconclusive.
Some of the problem is simply limited data: despite the fact that Wuhan has advanced virology labs, there is only patchy sampling of SARS-CoV-2 sequences from the first months of the city's explosive outbreak.
Other than a set of multiply sequenced samples collected in late December of 2019 from a dozen patients connected to the Huanan Seafood Market~\citep{WHO2021origins}, just a handful of Wuhan sequences are available from before late January of 2020 (see analysis in this study below).
This paucity of sequences could be due in part to an order that unauthorized Chinese labs destroy all coronavirus samples from early in the outbreak, reportedly for ``laboratory biological safety'' reasons~\citep{pinghui2020SCMP}.

@@

\subsection{The data are sufficient to determine the viral sequence from the start of spike through the end of ORF10 for some samples}
\citet{wang2020medRxiv} sequenced PCR amplicons covering nucleotide sites 21,563 to 29,674 of the SARS-CoV-2 genome, which spans from the start of the spike gene to the end of ORF10.
They also sequenced a short amplicon generated by nested PCR that covered a fragment of ORF1ab spanning sites $\sim$15,080 to 15,550.
In this paper, I only analyze the region from spike through ORF10 because this is a much longer contiguous sequence and the amplicons were generated by conventional rather than nested PCR.
I slightly trimmed the region of interest to 21,570 to 29,550 because many samples had poor coverage at the termini.

@@

I also determined the consensus sequence of the plasmid control used by \citet{wang2020medRxiv} from the recovered sequencing data, and found that it had mutations C28144T and G28085T relative to proCoV2, which means that in the region of interest this control matches Wuhan-Hu-1 with the addition of G28085T.
Since none of the viral samples in Table~\ref{tab:mutations} contain G28085T and the samples that prove most relevant below also lack C28144T (which is a frequent natural mutation among early Wuhan sequences), plasmid contamination did not afflict the viral samples in the deleted sequencing project.

@@

This conundrum is illustrated in Figure~\ref{fig:deltadist_RaTG13}, which plots the collection date of SARS-CoV-2 sequences in GISAID versus the relative number of mutational differences from RaTG13~\citep{zhou2020pneumonia}, which is the bat coronavirus with the highest full-genome sequence identity to SARS-CoV-2.
The earliest SARS-CoV-2 sequences were collected in Wuhan in December, but these sequences are more distant from RaTG13 than sequences collected in January from other locations in China or even other countries (Figure~\ref{fig:deltadist_RaTG13}).
The discrepancy is especially pronounced for sequences from patients who had visited the Huanan Seafood Market~\citep{WHO2021origins}.
All sequences associated with this market differ from RaTG13 by at least three more mutations than sequences subsequently collected at various other locations (Figure~\ref{fig:deltadist_RaTG13})---a fact that is difficult to reconcile with the idea that the market was the original location of spread of a bat coronavirus into humans.
Importantly, all these observations also hold true if SARS-CoV-2 is compared to other related bat coronaviruses~\citep{lytras2021exploring} such as RpYN06~\citep{zhou2021identification} or RmYN02~\citep{zhou2020novel} rather than RaTG13 (Figure~\ref{suppfig:deltadist_RpYN06_RmYN02}).

 \begin{figure*}
 \centerline{
 \includegraphics[width=0.31\linewidth, valign=t, clip=true, trim=0in 4.4in 0in 0in]{figures/tree_images/hCoV-19-USA-WA1-2020_RaTG13_without_deleted_seqs.pdf}
 \hspace{0.02\linewidth}
 \includegraphics[width=0.33\linewidth, valign=t, clip=true]{figures/tree_images/hCoV-19-Guangdong-HKU-SZ-002-2020_RaTG13_without_deleted_seqs.pdf}
  \hspace{0.02\linewidth}
 \includegraphics[width=0.31\linewidth, valign=t, clip=true, trim=0in 4.4in 0in 0in]{figures/tree_images/hCoV-19-Shandong-LY005-2-2020_RaTG13_without_deleted_seqs.pdf}
 }
 \caption{
 Phylogenetic trees of SARS-CoV-2 sequences in GISAID with multiple observations among viruses collected before Februrary, 2020.
 The trees are identical except they are rooted to make the progenitor each of the three sequences with highest identity to the RaTG13 bat coronavirus outgroup.
Nodes are shown as pie charts with areas proportional to the number of observations of that sequence, and colored by where the viruses were collected.
The mutations on each branch are labeled, with mutations towards the nucleotide identity in the outgroup in purple.
The labels at the top of each tree give the first known virus identical to each putative progenitor, as well as mutations in that progenitor relative to proCoV2~\citep{kumar2021evolutionary} and Wuhan-Hu-1.
The monophyletic group containing C28144T is collapsed into a node labeled ``clade B'' in concordance with the naming scheme of \citet{rambaut2020dynamic}; this clade contains Wuhan-Hu-1.
 Figure~\ref{suppfig:tree_RpYN06_RmYN02} shows identical results are obtained if the outgroup is RpYN06 or RmYN02.
\label{fig:tree_RaTG13}
 }
 \end{figure*}

This conundrum can be visualized in a phylogenetic context by rooting a tree of early SARS-CoV-2 sequences so that the progenitor sequence is closest to the bat coronavirus outgroup.
If we limit the analysis to sequences with at least two observations among strains collected no later than January 2020, there are three ways to root the tree in this fashion since there are three different sequences equally close to the outgroup (Figure~\ref{fig:tree_RaTG13}, Figure~\ref{suppfig:tree_RpYN06_RmYN02}).
Importantly, none of these rootings place any Huanan Seafood Market viruses (or other Wuhan viruses from December 2019) in the progenitor node---and only one of the rootings has any virus from Wuhan in the progenitor node (in the leftmost tree in Figure~\ref{fig:tree_RaTG13}, the progenitor node contains Wuhan/0126-C13/2020, which was reportedly collected on January 26, 2020).
Therefore, inferences about the progenitor of SARS-CoV-2 based on comparison to related bat viruses are inconsistent with other evidence suggesting the progenitor is an early virus from Wuhan~\citep{pipes2021assessing}.

@@

\begin{figure}
\centering
\includegraphics[width=0.75\linewidth]{figures/deltadist_jitter.png}
\caption{
Relative mutational distance from RaTG13 bat coronavirus outgroup calculated \emph{only} over the region of the SARS-CoV-2 genome covered by sequences from the deleted data set (21,570--29,550).
The plot shows sequences in GISAID collected before February of 2020, as well as the 13 early Wuhan epidemic sequences in Table~\ref{tab:mutations}.
Mutational distance is calculated relative to proCoV2, and points are jittered on the y-axis.
Go to {\color{red} add link} for an interactive version of this plot that enables toggling the outgroup to RpYN06 or RmYN02, mouseovers to see details for each point, and adjustment of jittering.
\label{fig:deltadist_jitter}
}
\end{figure}

 \begin{figure*}
 \centerline{
 \includegraphics[width=0.31\linewidth, valign=t, clip=true, trim=0in 6.75in 0in 0in]{figures/tree_images/hCoV-19-USA-WA1-2020_RaTG13_with_deleted_seqs.pdf}
 \hspace{0.02\linewidth}
 \includegraphics[width=0.33\linewidth, valign=t, clip=true]{figures/tree_images/hCoV-19-Guangdong-HKU-SZ-002-2020_RaTG13_with_deleted_seqs.pdf}
  \hspace{0.02\linewidth}
 \includegraphics[width=0.31\linewidth, valign=t, clip=true, trim=0in 6.75in 0in 0in]{figures/tree_images/hCoV-19-Shandong-LY005-2-2020_RaTG13_with_deleted_seqs.pdf}
 }
 \caption{
Phylogenetic trees like those in Figure~\ref{fig:tree_RaTG13} with the addition of the early Wuhan epidemic sequences from the deleted data set, and Guangdong patients infected in Wuhan prior to January 5 annotated separately.
Because the deleted sequences are partial, they cannot all be placed unambiguously on the tree.
Therefore, they are added to each compatible node proportional to the number of sequences already in that node.
The deleted sequences with C28144T (clade B) or C29095T (putative progenitor in middle tree) can be placed relatively unambiguously as defining mutations occur in the sequenced region, but those that lack either of these mutations are compatible with a large number of nodes including the proCoV2 putative progenitor.
 Figure~\ref{suppfig:tree_RpYN06_RmYN02} demonstrates that the results are identical if RpYN06 or RmYN02 is instead used as the outgroup.
\label{fig:tree_with_deleted}
 }
 \end{figure*}

Repeating the analysis of the previous section with these changes shows that several sequences from the deleted project and all sequences from patients infected in Wuhan but sequenced in Guangdong are more similar to the bat coronavirus outgroup than sequences from the Huanan Seafood Market (Figure~\ref{fig:deltadist_jitter}).
This fact suggests that the market sequences, which are the primary focus of the genomic epidemiology in the joint WHO-China report~\citep{WHO2021origins}, are not representative of the viruses that were circulating in Wuhan in late December of 2019 and early January of 2020.

@@

We can also qualitatively examine the three progenitor placements in Figure~\ref{fig:tree_with_deleted} using the principle employed by \citet{worobey2020emergence} to help evaluate scenarios for emergence of SARS-CoV-2 in Europe and North America: namely that during a growing outbreak, a progenitor is likely to give rise to multiple branching lineages.
This principle is especially likely to hold for the scenarios in Figure~\ref{fig:tree_with_deleted}, since there are multiple individuals infected with each putative progenitor sequence, implying multiple opportunities to transmit descendants with new mutations.
Using this qualitative principle, the middle scenario in Figure~\ref{fig:tree_with_deleted} seems most plausible, the leftmost (proCoV2) scenario also seems plausible, and the rightmost scenario seems less plausible.
I acknowledge these arguments are purely qualitative and lack the formal statistical analysis of \citet{worobey2020emergence}---but as discussed below, there may be wisdom in qualitative reasoning when there are valid concerns about the nature of the underlying data.

\section{Discussion}
I have identified and recovered a deleted data set of partial SARS-CoV-2 sequences from the early Wuhan epidemic.
Analysis of these sequences leads to several important conclusions.
First, the Huanan Seafood Market sequences that were the focus of the joint WHO-China report~\citep{WHO2021origins} are clearly not representative of all SARS-CoV-2 in Wuhan early in the epidemic.
Instead, the deleted data set as well as analysis of existing sequences from Wuhan-infected patients hospitalized in Guangdong shows early Wuhan sequences often carried the T29095C mutation and were less likely to carry T8782C / C28144T than the sequences in the joint WHO-China report~\citep{WHO2021origins}.
Second, given current data, there are two plausible identities for the progenitor of all known SARS-CoV-2 sequences.
One of these putative progenitors is proCoV2 described by \citet{kumar2021evolutionary}, and the other is a sequence that carries three mutations (C8782T, T28144C, and C29095T) relative to Wuhan-Hu-1.
Crucially, both putative progenitors are three mutations closer to SARS-CoV-2's bat coronavirus relatives than sequences from the Huanan Seafood Market.
Note also that the progenitor of all currently known SARS-CoV-2 sequences could still be downstream of the sequence that actually infected patient zero depending on the transmission dynamics of the earliest infections.

In addition, the fact that such an informative data set was deleted carries implications beyond those gleaned directly from the recovered sequences.
Samples from early outpatients in the Wuhan epidemic are a gold mine for anyone seeking to understand the early spread of the virus.
Even my analysis of the partial sequences is revealing, and it clearly would have been more scientifically informative to fully sequence the samples rather than surreptitiously delete the partial sequences.
There is no plausible scientific reason for the deletion: the sequences are perfectly concordant with the samples described in \citet{wang2020medRxiv,wang2020small}, there are no corrections to the paper, the paper states proper human subjects approval was obtained, and the sequencing data show no evidence of plasmid or sample-to-sample contamination.
It therefore seems likely that the data were deleted from the SRA to obscure their existence from scientists searching such databases.

There are several caveats...
Sequences only partial and lacking full metadata...
Something about lacking formal analyses of some other studies, but maybe this is not so bad.
Still not parsimonious

Finally, implications for further study. getting info from NIH on these and others.
